### 2025-01-24 skripsi/chapter1.tex
move operator learning to related works, update revision list, mention mathematical  guarantee of conventional numerical methods in comparison scenario

\section{Contribution}
\noindent In achieving the aims of this study the following contributions are made:
\begin{enumerate}
    \item A novel use LSSVR which has a convex objective to learn solution operators of partial differential equations and operators in general.
    \item Faster time to solution compared to conventional methods provided by widely used libraries like SciPy.
    \item Interpretable machine learning model trained on operator data.
\end{enumerate}

>>>>

\section{Contribution}
\noindent In achieving the aims of this study the following contributions are made:
\begin{enumerate}
    \item A novel use LSSVR which has a convex objective to learn solution operators of partial differential equations and operators in general.
    \item Faster time from parameter to solution compared to conventional methods provided by widely used libraries like SciPy.
    \item Interpretable data-driven model for the mapping between spectral coefficients.
\end{enumerate}

### 2025-01-24 skripsi/chapter2.tex
add discussion, add comparison, change conclusion, revise future work
move operator learning to related works, update revision list, mention mathematical  guarantee of conventional numerical methods in comparison scenario

\section{Operator Learning}
\noindent Machine learning, as with many other fields, has found its foothold in scientific computing. The term that has become popular for this field is Scientific Machine Learning (SciML) \autocite{bakerWorkshopReportBasic2019}. This field, just like scientific computing, is interdisciplinary with mathematics, physics, computer science, and other disciplines contributing to solve the specific challenges involved with datasets specific to these individual disciplines. The field seeks to extract insights from scientific datasets by building new methods that are scalable, interpretable, and reliable. This is partly driven by the advances in computing, specifically machine learning, and the potential for and accumulation of scientific data. The motivations for developing domain-specific machine learning tools include the difficulty in using and interpreting general purpose ML algorithms for the problems researchers seek to solve with scientific datasets.

@@

There are currently different model architectures for the learning operators. Most of these are implemented as Neural Networks (NN). The Neural Operator proposes a method to generalize NNs to learning mappings between function spaces \autocite{li2021fourier}. The Fourier Neural Operator (FNO) variant adds Fourier transforms to the composition of layers within the NN\@. Another NN layer is added to learn the mapping within the Fourier domain. The outputs of this layer is then Inverse transformed and passed on. This entire process is called the FNO block. Adding multiple of these blocks allow for learning more complex relationships. Another approach called Spectral Neural Operators, instead maps the coefficients of inputs to the coefficients of output functions \autocite{fanaskovSpectralNeuralOperators2023}. For coefficients obtained from Fourier transforms, the model is approximating the standard spectral method by taking advantage of the properties of the Fourier coefficients we discuss in \lccref{sec:fourier_discussion}. Essentially, derivatives are multiplication in the Fourier domain. A similar work extended the idea by leveraging the known PDE to be part of the loss function \autocite{du2024neural}. The model is penalized for the residual computed using the PDE\@. This allows the NN to be trained in a self-supervised setting. Another major model is the so called DeepONet architecture proposed by \textcite{luLearningNonlinearOperators2021}. This model can be seen as foregoing the basis function the Fourier Transform enforces and instead learns the basis functions from the data \autocite{meurisMachinelearningbasedSpectralMethods2023}. Because of this, the model has two components. The first, named the trunk network acts as the learned basis functions. This network is given evaluation coordinates and computes the values of the \enquote{basis functions}. The second network, called the branch network, learns the mapping between the input function representation and the coefficients for the output function. The dot product of the outputs of both trunk and branch networks are evaluated to obtain the value of the predicted function at the specified coordinate. The final proposed approach we will discuss here doesn't use Deep Learning. The approach proposed by \textcite{nelsenOperatorLearningUsing2024} uses Random Feature Models (RFM) to learn operators. This model enjoys the advantage of using one of the simplest forms of machine learning.

@@

To compute the optimal parameters, first the point at which the gradient is zero for each parameter must be found. This is simply by equating to zero the derivative of the Lagrangian with respect to each parameter. This is called the Karush-Khun-Tucker optimality condition. Computing this for our Lagrangian results in \lccref{eq:lssvr_KKT}.
\begin{equation}
  \begin{split}
    \pdv{L}{W} = 0 \rightarrow          & W           = & \sum_{k=1}^{n}\alpha_{k}\vb{x}_{k}                      \\
    \pdv{L}{b} = 0 \rightarrow          & 0           = & \sum_{k=1}^{n}\alpha_{k}                                \\
    \pdv{L}{e_{k}} = 0 \rightarrow      & \alpha_{k}  = & C e_{k}                            & \qquad k=1,\dots,n \\
    \pdv{L}{\alpha_{k}} = 0 \rightarrow & y_{k}       = & W^{\intercal}\vb{x}_{k}+b+e_{k}    & \qquad k=1,\dots,n \\
  \end{split}\label{eq:lssvr_KKT}
\end{equation}

@@

The second tool is the p-vector. This tool seeks to inform the learned contributions of each input feature towards the output. The Lagrangian multipliers contain information on the contribution of each training sample towards the output. To link this with the input features, the multipliers are multiplied against the matrix of training inputs. This is done by transposing the matrix of training inputs and performing a matrix multiplication against the Lagrange multipliers. The result shows how much contribution each input contributes to the output. If we generalize to multiple outputs, the result is a matrix with the rows representing input features and columns representing each output.

@@

% Advantages and Disadvantages
LSSVMs have many benefits but also its own trade-offs. The main disadvantage of LSSVMs is the fact that they do not have the sparse property of SVMs. This can result in more computation during training and prediction since every single sample would need to be included in the computations. A simple mitigation can be done by filtering training samples with small absolute values of Lagrangian multipliers \autocite{haifengwangComparisonSVMLSSVM2005}. However, the benefits of simple matrix operations outweigh the disadvantages. This is because of the parallelization benefits of algorithms for simple matrix operations. With access to parallel hardware accelerators, the speedup for both training and inference can be very large.

>>>>

\section{Related Works}
\noindent Machine learning, as with many other fields, has found its foothold in scientific computing. The term that has become popular for this field is Scientific Machine Learning (SciML) \autocite{bakerWorkshopReportBasic2019}. This field, just like scientific computing, is interdisciplinary with mathematics, physics, computer science, and other disciplines contributing to solve the specific challenges involved with datasets specific to these individual disciplines. The field seeks to extract insights from scientific datasets by building new methods that are scalable, interpretable, and reliable. This is partly driven by the advances in computing, specifically machine learning, and the potential for and accumulation of scientific data. The motivations for developing domain-specific machine learning tools include the difficulty in using and interpreting general purpose ML algorithms for the problems researchers seek to solve with scientific datasets.

@@

There are currently different model architectures for the learning operators. Most of these are implemented as Neural Networks (NN). The Neural Operator proposes a method to generalize NNs to learning mappings between function spaces \autocite{li2021fourier}. The Fourier Neural Operator (FNO) variant adds Fourier transforms to the composition of layers within the NN\@. Another NN layer is added to learn the mapping within the Fourier domain. The outputs of this layer is then Inverse transformed and passed on. This entire process is called the FNO block. Adding multiple of these blocks allow for learning more complex relationships. Another approach called Spectral Neural Operators, instead maps the coefficients of inputs to the coefficients of output functions \autocite{fanaskovSpectralNeuralOperators2023}. For coefficients obtained from Fourier transforms, the model is approximating the standard spectral method by taking advantage of the properties of the Fourier coefficients we discuss in \lccref{sec:fourier_discussion}. Essentially, derivatives are multiplication in the Fourier domain. A similar work extended the idea by leveraging the known PDE to be part of the loss function \autocite{du2024neural}. The model is penalized for the residual computed using the PDE\@. This allows the NN to be trained in a self-supervised setting.

Another paradigm is the so called DeepONet architecture proposed by \textcite{luLearningNonlinearOperators2021}. This model can be seen as foregoing the basis function the Fourier Transform enforces and instead learns the basis functions from the data \autocite{meurisMachinelearningbasedSpectralMethods2023}. Because of this, the model has two component networks. The first, named the trunk network acts as the learned basis functions. This network is given evaluation coordinates and computes the values of the \enquote{basis functions} at said coordinates. The second network, called the branch network, learns the mapping between the input function representation and the coefficients for the output function. The dot product of the outputs of both trunk and branch networks are evaluated to obtain the value of the predicted function at the specified coordinate. The final proposed approach we will discuss here doesn't use Deep Learning. The approach proposed by \textcite{nelsenOperatorLearningUsing2024} uses Random Feature Models (RFM) to learn operators. This model enjoys the advantage of using one of the simplest forms of machine learning.

The literature also further into the learned basis functions of DeepONet. In their study, \textcite{meurisMachinelearningbasedSpectralMethods2023} utilize the trained trunk network to extract custom basis functions. These can then be used in a spectral procedure like Fourier basis functions. These have been found to be more accurate than the DeepONet by itself. Building on a two-stage training process that was originally done to improve performance of DeepONets \autocite{leeTrainingGeneralizationDeep2024}, \textcite{peyvanRiemannONetsInterpretableNeural2024} uses the two-stage method to obtain the basis functions for their proposed RiemannONet. The basis functions that are obtained are also investigated. The authors also reveal the contribution of the hidden layers towards the final shape of the basis functions using Singular Value Decomposition (SVD). This interpretation allows the authors to see how different layer compositions affect the contributions toward the basis functions. These works extend the understanding of learned basis functions. However, to our knowledge at the time of writing, there has been no discussion on interpreting the learned mapping between spectral coefficients. The interpretability of this is crucial to better understand what the model actually learned in terms of the operator between solution and parameter functions.

@@

To compute the optimal parameters, first the point at which the gradient is zero for each parameter must be found. This is simply by equating to zero the derivative of the Lagrangian with respect to each parameter. This is called the Karush-Khun-Tucker optimality condition. Computing this for our Lagrangian results in \lccref{eq:lssvr_kkt}.
\begin{subequations}
  \begin{align}
    \pdv{L}{W} = 0 \rightarrow          & W           = & \sum_{k=1}^{n}\alpha_{k}\vb{x}_{k}                      \label{eq:lssvr_kkt_w}                                             \\
    \pdv{L}{b} = 0 \rightarrow          & 0           = & \sum_{k=1}^{n}\alpha_{k}                                \label{eq:lssvr_kkt_b}                                             \\
    \pdv{L}{e_{k}} = 0 \rightarrow      & \alpha_{k}  = & C e_{k}                                                                        & \qquad k=1,\dots,n \label{eq:lssvr_kkt_e} \\
    \pdv{L}{\alpha_{k}} = 0 \rightarrow & y_{k}       = & W^{\intercal}\vb{x}_{k}+b+e_{k}                                                & \qquad k=1,\dots,n \label{eq:lssvr_kkt_a}
  \end{align}\label{eq:lssvr_kkt}
\end{subequations}

@@

The second tool is the p-vector. This tool seeks to inform the learned contributions of each input feature towards the output. The Lagrangian multipliers contain information on the contribution of each training sample towards the output. To link this with the input features, the multipliers are multiplied against the matrix of training inputs. This is done by transposing the matrix of training inputs and performing a matrix multiplication against the Lagrange multipliers. The result shows how much contribution each input contributes to the output. If we generalize to multiple outputs, the result is a matrix with the rows representing input features and columns representing each output. For the linear kernel, this would essentially be the weight vector as shown in \lccref{eq:lssvr_kkt_w}. However, the relationship is more abstract when we apply to nonlinear kernels because the weight vector is then the Lagrangian multipliers multiplied by the inputs in feature space (higher dimension). This means if we multiply the Lagrangian multipliers with the inputs in its original form, the result we get a projection of the contributions of each feature on the output.

@@

% Advantages and Disadvantages
LSSVMs have many benefits but also its own trade-offs. The main disadvantage of LSSVMs is the fact that they do not have the sparse property of SVMs. This can result in more computation during training and prediction since every single sample would need to be included in the computations. A simple mitigation can be done by filtering training samples with small absolute values of Lagrangian multipliers \autocite{haifengwangComparisonSVMLSSVM2005}. Larger training sets would also mean that the model would need to compute through the kernel matrix which scales as \(n^2\). This means that for larger datasets, optimization strategies such as only computing relevant parts of the kernel matrix would become necessary.

However, the benefits of using the LSSVM, such as simple matrix operations, outweigh the disadvantages. Specifically for the LSSVM compared to SVM with SMO, the model benefits from parallelization of algorithms for simple matrix operations. With access to parallel hardware accelerators, the speedup for both training and inference can be very large. Another benefit is the clear role of the parameters of the model \autocite{suykensLeastSquaresSupport2005}. The Lagrangian multiplier represents the influence of each sample on the model output. The bias term on the other hand represents the offset of the model output. This is in contrast to the weight parameters of deep neural networks which can represent any number of influence on the output. This fact makes SVMs much more clear to interpret in comparison to NNs \autocite{zhouInterpretingDeepVisual2019}. Against methods like kernel regression, SVMs enjoy the benefit of the training process being able to figure out which variables matter. In contrast, kernel regression would be less optimal since the regularization term in SVMs penalizes it for undue complexity.

### 2025-01-24 skripsi/chapter4.tex
move operator learning to related works, update revision list, mention mathematical  guarantee of conventional numerical methods in comparison scenario

The second scenario extends the goal of the first scenario by examining the model and its behavior when learning a nonlinear PDE\@. The data subsets with different viscosity values are intended to allow us to learn more about how this will affect the performance on the exact solutions for lower viscosity numbers. The model will learn to predict future states of systems governed by the Burgers' equation from past states. The model for each viscosity will be used to predict \lccref{eq:burgers_exact_solution} from initial conditions at time equals zero and the corresponding forcing term of a constant value of zero. The exact equation prediction will be done using a initial value problem formulation, also described as a rollout. The initial solution value along with the forcing term will be given to the model to predict the solution at the first time step. The model will then use the predicted solution with the forcing term to predict the solution at the second time step. This is repeated until the stopping condition of time \(t=10\). The forcing term for the exact solution is a constant zero. In addition to the rollout of the exact solution, we also perform rollout on of the test set. This will be done in comparison to traditional numerical methods, namely the commonly used method of lines in time using SciPy and torchdiffeq and finite differences in space with NumPy gradient and PyTorch gradient respectively. Specifically, the SciPy odeint function uses lsoda from the FORTRAN library odepack. And, the torchdiffeq version will use Implicit Adams-Bashforth-Moulton (IABM) method. Both of these were chosen for their availability for use in Python and therefore the kind of numerical solver one could set up with relative reliability.

>>>>

The second scenario extends the goal of the first scenario by examining the model and its behavior when learning a nonlinear PDE\@. The data subsets with different viscosity values are intended to allow us to learn more about how this will affect the performance on the exact solutions for lower viscosity numbers. The model will learn to predict future states of systems governed by the Burgers' equation from past states. The model for each viscosity will be used to predict \lccref{eq:burgers_exact_solution} from initial conditions at time equals zero and the corresponding forcing term of a constant value of zero. The exact equation prediction will be done using a initial value problem formulation, also described as a rollout. The initial solution value along with the forcing term will be given to the model to predict the solution at the first time step. The model will then use the predicted solution with the forcing term to predict the solution at the second time step. This is repeated until the stopping condition of time \(t=10\). The forcing term for the exact solution is a constant zero. In addition to the rollout of the exact solution, we also perform rollout on of the test set. This will be done in comparison to traditional numerical methods, namely the commonly used method of lines in time using SciPy \autocite{virtanenSciPy10Fundamental2020} and torchdiffeq \autocite{Chen_torchdiffeq_2021} with finite differences in space using NumPy gradient and PyTorch gradient respectively. Specifically, the SciPy odeint function uses lsoda from the FORTRAN library ODEPACK \autocite{osti_145724}. And, the torchdiffeq version will use Implicit Adams-Bashforth-Moulton (IABM) method \autocite{bashforth1883attempt,moulton1926new}. Both of these were chosen for their availability for use in Python and therefore the kind of numerical solver one could set up with relative reliability. The choice of conventional physics based numerical methods is due to its mathematical guarantee that machine learning models do not possess \autocite{schiesserNumericalMethodLines2012,karniadakisPhysicsinformedMachineLearning2021}.
